---
title: "Quarto Cheatsheet"
author: "Hippolyte Menou"
---

In Quarto, code blocks can be customized using various parameters to control their behavior and appearance. These parameters are specified using the `#|` syntax within the code block. Here are some of the key parameters you can use:

1. **`label`**: A label for the code block, which can be used for referencing.

```{r}
#| label: fig-theme-setting
```

2. **`echo`**: Controls whether the code itself is displayed in the output.

```{r}
#| echo: false
```

3. **`include`**: Controls whether the code and its output are included in the final document.

```{r}
#| include: true
```

4. **`eval`**: Controls whether the code is evaluated.

```{r}
#| eval: false
```

5. **`results`**: Controls how the results of the code are displayed (e.g., `"asis"`, `"hold"`, `"hide"`, `"markup"`).

```{r}
#| results: "asis"
```

6. **`warning`** and **`message`**: Controls whether warnings and messages are displayed.

```{r}
#| warning: false
#| message: false
```

7. **`error`**: Controls whether errors are displayed.

```{r}
#| error: true
```

8. **`fig.cap`**: A caption for the figure.

```{r}
#| fig.cap: "Air Quality"
```

9. **`fig.height`** and **`fig.width`**: Controls the height and width of the figure.

```{r}
#| fig.height: 6
#| fig.width: 8
```

10. **`fig.align`**: Controls the alignment of the figure (e.g., `"left"`, `"center"`, `"right"`).

```{r}
#| fig.align: "center"
```

11. **`out.width`** and **`out.height`**: Controls the width and height of the output.

```{r}
#| out.width: "50%"
#| out.height: "300px"
```

12. **`cache`**: Controls whether the results of the code block are cached.

```{r}
#| cache: true
```

13. **`collapse`**: Controls whether the code and output are collapsed into a single block.

```{r}
#| collapse: true
```

14. **`code-fold`**: Controls whether the code is folded (hidden) by default.

```{r}
#| code-fold: true
```

15. **`code-summary`**: Provides a summary of the code when it is folded.

```{r}
#| code-summary: "This code generates a plot."
```

16. **`code-line-numbers`**: Controls whether line numbers are shown in the code block.

```{r}
#| code-line-numbers: true
```

17. **`code-tools`**: Controls whether code tools (e.g., copy button) are shown.

```{r}
#| code-tools: true
```

18. **`code-overflow`**: Controls how code overflow is handled (e.g., `"scroll"`, `"wrap"`).

```{r}
#| code-overflow: "scroll"
```

19. **`code-link`**: Enables hyper-linking of functions within code blocks to their online documentation.

```{r}
#| code-link: true
```

20. **`layout-ncol`**: Controls the number of columns in a layout for figures.

```{r}
#| layout-ncol: 2
```

21. **`layout-nrow`**: Controls the number of rows in a layout for figures.

```{r}
#| layout-nrow: 2
```

22. **`df-print`**: Controls the default data frame printing method (e.g., `"default"`, `"kable"`, `"tibble"`, `"paged"`).

```{r}
#| df-print: "kable"
```

