---
title: "2018-04-03 - US Tuition Costs"
author: "Hippolyte Menou"
format: html
editor: visual
---

# Introduction

This analysis explores the evolution of university tuition costs across different states in the United States over a twelve-year period (2004-2016). We'll investigate patterns in tuition increases, state-by-state variations, and potential relationships between initial costs and growth rates.

# Data Preparation

We begin by loading necessary libraries for data manipulation, visualization, and formatting. Our visual theme is minimalist, using a custom color palette and the Fira Sans font family. We've also set up currency formatting for clearer presentation of dollar amounts.

```{r Libraries}
library(glue)
library(scales)
library(ggtext)
library(ggdist)
library(readxl)
library(paletteer)
library(patchwork)
library(tidyverse)
```

```{r Theme}
theme_set(theme_minimal())

theme_update(
  legend.position = "None"
)

font_family <- "Fira Sans"

palette <- c("#D3F2A3FF", "#97E196FF", "#6CC08BFF", "#4C9B82FF", "#217A79FF", "#105965FF", "#074050FF")


label_dollars <- label_currency(accuracy = 1)

```

The dataset contains average tuition costs by state from 2004-05 to 2015-16 academic years. We reshape the data into a longer format to facilitate various types of analysis.

```{r Import Data}
us_avg_tuition <- read_xlsx("us_avg_tuition.xlsx")

us_avg_tuition_longer <- us_avg_tuition %>% 
  pivot_longer(`2004-05`:`2015-16`, names_to = "Season", values_to = "avg_tuition") 

us_avg_tuition_longer %>% 
  glimpse()
```

# How did tuition costs increased over time ?

Our analysis reveals several interesting patterns in tuition cost evolution:

1.  **Overall Growth Patterns**: We categorize states into six groups based on their tuition increase rates:

    -   Decrease : Only one state had an decrease in the average tuition costs (Ohio)
    -   Slow increase (0-23%)
    -   Below median increase (23-50%)
    -   Above median increase (50-75%)
    -   Fast increase (75-100%)
    -   More than double increase (\>100%) : Multiple states had more than doubled

2.  **Distribution Changes**: The visualization shows how tuition costs have become more concentrated around the median over time, with some notable exceptions at the upper end. The median tuition increased from \${label_dollars(median_tuition_2005)} in 2004-05 to \${label_dollars(median_tuition_2015)} in 2015-16.

3.  **Consistency of Increases**: Interestingly, despite the 2008 financial crisis, tuition increases remained relatively consistent, with continued growth resuming immediately in the following year.

```{r}
us_avg_tuition %>% 
  mutate(difference = `2015-16` - `2004-05`) %>% 
  ggplot(aes(x = fct_reorder(State, difference))) + 
  geom_col(aes(y = `2015-16`), fill = "red") + 
  geom_col(aes(y = `2004-05`), fill = "orange") + 
  geom_col(aes(y = - difference)) +
  coord_flip()

diff_cutoff <- c(0, 0.23, 0.5, 0.75, 1)
diff_cutoff_names <- c("Decrease", "Slow increase", "Below median increase", "Above median increase", "Fast increase", "More than double increase")

us_avg_tuition %>% 
  mutate(difference = `2015-16` - `2004-05`,
         diff_perc = round(difference / `2004-05`, 2),
         diff_group = cut(
           diff_perc, 
           breaks = c(-Inf, diff_cutoff, Inf), 
           labels = diff_cutoff_names,
           right = FALSE),
         diff_group = factor(diff_group, levels=diff_cutoff_names)) %>% 
  pivot_longer(`2004-05`:`2015-16`, names_to = "Season", values_to = "avg_tuition")  %>% 
  filter(Season %in% c("2004-05", "2015-16")) %>% 
  
  ggplot(aes(x = Season, avg_tuition, group=State)) + 
  geom_point(aes(color=Season)) + 
  geom_line() + 
  facet_wrap(diff_group~., nrow=1)


us_avg_tuition %>% 
  mutate(difference = `2015-16` - `2004-05`) %>% 
  ggplot(aes(x = `2004-05`, y = `2015-16`)) + 
  geom_segment(aes(x = `2004-05`, y = `2004-05`, xend = `2004-05`, yend = `2015-16`, colour = "segment"), size=1, alpha=0.8) + 
  geom_point(size=4, alpha=0.5) + 
  geom_abline(slope = 1, intercept = 0, size=2) + 
  
  lims(x= c(0 , NA), y = c(0 , NA))

# us_avg_tuition %>% 
#   mutate(difference = `2015-16` - `2004-05`,
#          diff_perc = round(difference / `2004-05`, 2)) %>% 
#   ggplot(aes(group=State)) + 
#     geom_point(aes(x="A", y=0)) + 
#     geom_point(aes(x = "B", y=diff_perc))

median_tuition <- us_avg_tuition_longer %>% 
  group_by(Season) %>% 
  summarize(median = median(avg_tuition)) %>% 
  ungroup() 

median_tuition_2005 <- median_tuition %>% filter(Season == "2004-05") %>% head(1) %>% pull(median)
median_tuition_2015 <- median_tuition %>% filter(Season == "2015-16") %>% head(1) %>% pull(median)

us_avg_tuition %>% 
  pivot_longer(`2004-05`:`2015-16`, names_to = "Season", values_to = "avg_tuition")  %>% 
  
  ggplot(aes(x = avg_tuition, y=Season)) + 
  geom_vline(xintercept = median_tuition_2005, col = "grey30", lty = "dashed") +
  annotate("text", y = 0.5, x = median_tuition_2005 + 50, label = "Median tuition in 2004/05", size = 3, hjust = 0) +
  annotate("text", y = 0.5, x = median_tuition_2005 - 50, label = label_dollars(median_tuition_2005), size = 3, hjust = 1) +
  
  
  geom_vline(xintercept = median_tuition_2015, col = "grey30", lty = "dashed") +
  annotate("text", y = 0.5, x = median_tuition_2015 + 50, label = "Median tuition in 2015/16", size = 3, hjust = 0) +
  annotate("text", y = 0.5, x = median_tuition_2015 - 50, label = label_dollars(median_tuition_2015),
           size = 3, hjust = 1) +
  
  stat_halfeye(fill_type = "segments", alpha = 0.3) +
  stat_interval() +
  stat_summary(geom = "point", fun = median) +
  
  scale_y_discrete(expand = expansion(add = c(1, 0))) +
  scale_x_continuous(labels = scales::dollar_format()) +
  
  scale_color_manual(values = palette)



```

The "lagging university increased faster to get more concentrated around the mean. A few escaped to reach the top

No increase in 2008 yet the trend is consistent as soon as the following year

# Average tuition cost trends over time ?

```{r}
```

# Did cheaper schools made their price increase faster ?

```{r}
```

```{r}
```

```{r}
```
