---
title: "European Drug Development"
author: "Hippolyte Menou"
format: html
editor: visual
---

# Introduction

This analysis explores drug development and approval patterns in Europe, focusing particularly on orphan drugs - medicines developed specifically for rare diseases. The data comes from the European Medicines Agency (EMA) via Miquel Anglada Girotto on GitHub. Miquel wrote about his exploration of the data.

We used the source table of all EPARs for human and veterinary medicines, rather than Miquel's scraped data. It includes information about drug authorizations, therapeutic areas, and timelines from 1995 to 2022.

# Data Preparation

After loading necessary visualization and data manipulation libraries, we import the dataset from TidyTuesday's repository. The data originates from the European Medicines Agency's table of all EPARs (European Public Assessment Reports) for human and veterinary medicines.

```{r Libraries}
# devtools::install_github("teunbrand/elementalist")

library(scales)
library(ggridges)
library(tidyverse)
library(paletteer)
library(patchwork)
library(elementalist)  
```

```{r Import data}
drugs <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2023/2023-03-14/drugs.csv')

drugs %>% 
  glimpse
```

# Analysis of Drug Categories

Looking at the therapeutic areas for human drugs with authorized status, we can see the distribution of drug development across different medical fields. This helps us understand where pharmaceutical research and development efforts have been concentrated.

```{r Drug classes}
drugs %>% 
  mutate(category=="human", authorisation_status=="authorized") %>% 
  count(therapeutic_area, sort=T)
```

# Time to Market Analysis

We examine the time between opinion date and decision date for non-generic, non-biosimilar human drugs. The violin plot comparison between orphan and non-orphan medicines reveals the variation in approval timelines, highlighting potential differences in the review process for these categories.

```{r}
drugs %>% 
  mutate(time_to_market = decision_date - date_of_opinion) %>% 
  filter(category=="human", !generic, !biosimilar, time_to_market>0) %>% 
  ggplot(aes(x = time_to_market, y = orphan_medicine)) + 
    geom_violin()
```

# Deep Dive into Orphan Drugs

For our main analysis, we create two complementary visualizations using Jack Bush's color palette to explore orphan drug development over time:

## Temporal Analysis (1995-2022): 

-   Shows the yearly number of drug evaluations by the EMA
-   Highlights the impact of EU orphan regulation (effective 2000)
-   Demonstrates the relatively slow initial uptake of orphan drug submissions
-   Notes the recent trend, with only 20 orphan drugs submitted in 2022

```{r}

jack_bush_palette <- paletteer_d("lisa::JackBush_1")
names(jack_bush_palette) <- c("Drug authorised", "Drug withdrawn", "Orphan drug authorised", "Orphan drug withdrawn")
jack_bush_palette

theme_set(theme_minimal())
theme_update(
  # Legend Styling
  legend.title = element_blank(), 
  legend.position = "none", 
  
  # Title Styling
  plot.title = element_text(hjust = 0.5, face = "bold", size = 16),
  plot.subtitle = element_text(hjust = 0.5),
  
  
  #
  panel.grid.minor = element_blank(),
  
  # Margins
  plot.margin = margin(t = 20, r = 20, b = 20, l = 20),
)

```

```{r}
orphan_drugs_years <- drugs %>% 
  mutate(year = year(date_of_opinion)) %>%  # date_of_opinion, decision_date, first_published, marketing_authorisation_date, revision_date
  filter(!is.na(year)) %>% 
  group_by(orphan_medicine, authorisation_status) %>% 
  count(year) %>% 
  mutate(authorisation_status = if_else(authorisation_status=="authorised", T, F),
         approved_orphan = case_when(
           orphan_medicine & authorisation_status ~ "Orphan drug authorised",  # TRUE/TRUE
           orphan_medicine & !authorisation_status ~ "Orphan drug withdrawn",  # TRUE/FALSE
           !orphan_medicine & authorisation_status ~ "Drug authorised",   # FALSE/TRUE
           !orphan_medicine & !authorisation_status ~ "Drug withdrawn"   # FALSE/FALSE
         ) %>% factor(levels=c("Drug withdrawn", "Drug authorised", "Orphan drug withdrawn", "Orphan drug authorised")))  

orphan_drugs_years
```

```{r}
orphan_years_plot <- orphan_drugs_years %>% 
  ggplot(aes(x=year, y=n, fill=approved_orphan)) + 
  geom_col() + 
  
  scale_fill_manual(values=jack_bush_palette, breaks=c("Drug authorised", "Drug withdrawn", "Orphan drug authorised", "Orphan drug withdrawn")) + 
  
  scale_x_continuous(breaks=seq(1995, 2020, 5)) +
  scale_y_continuous(breaks=seq(0, 120, 20)) +

  # Number below horizon
  geom_text(data = orphan_drugs_years %>% filter(approved_orphan == "Orphan drug authorised"), aes(label=n), colour="#F45F40FF", y = -2, fontface="bold") +
  
  labs(x = "Year of Opinion",
       y = "Number of Drugs", 
       title = "Number of drugs evaluated by the \n European Medicines Agency each year",
       subtitle = "Submissions for orphan designation lagged after the EU orphan regulation became \neffective in 2000. In 2022, only 20 orphan drugs were submitted for evaluation.",
       # caption = "Data source: European Medicines Agency",
  ) +
  
  theme(
    # Legend Styling
    legend.position = "bottom", 
    legend.box.background = element_rect_round(fill = "white", colour = "black"),
  )

orphan_years_plot
```

##  Overall Statistics: 

-   Orphan drugs represent 12% of all approved drugs
-   Both orphan and regular drugs maintain similar authorization rates of around 90%
-   Provides a clear comparison of total volumes between regular and orphan drugs

```{r}
orphan_drugs_summary <- orphan_drugs_years %>% 
  group_by(orphan_medicine, approved_orphan) %>% 
  summarise(total = sum(n)) %>% 
  ungroup %>% group_by(orphan_medicine) %>% add_tally(total) %>% ungroup %>% 
  mutate(perc = round(total / n, 2), perc2 = round(n / sum(total), 2))

orphan_drugs_summary

orphan_total_plot <- orphan_drugs_summary %>% 
  ggplot(aes(x = orphan_medicine, y = total, fill = approved_orphan)) + 
  geom_col() + 
  
  # Number below the horizon
  geom_text(data = orphan_drugs_summary %>% filter(approved_orphan %in% c("Orphan drug authorised", "Drug authorised")), aes(label=n), y=-20, fontface="bold", colour="black") + 
  
  # Percentage of approval
  geom_text(data = orphan_drugs_summary %>% filter(approved_orphan %in% c("Orphan drug authorised", "Drug authorised")), aes(label=percent(perc)), position = position_stack(vjust = .5), colour="darkgray", fontface="bold") + 
  
  # Percentage of orphan drugs total
  geom_text(data = orphan_drugs_summary %>% filter(approved_orphan %in% c("Orphan drug authorised", "Drug authorised")), aes(label=percent(perc2)), y=1100, fontface="bold", size=6) + 
  
  scale_fill_manual(values=jack_bush_palette, breaks=c("Drug authorised", "Drug withdrawn", "Orphan drug authorised", "Orphan drug withdrawn")) +
  scale_x_discrete(labels = c("FALSE" = "Drugs", "TRUE" = "Orphan drugs")) +  # rename orphan drugs T/F
  scale_y_continuous(breaks = seq(0, 1200, 200), expand = expansion(mult = c(0.05, 0.15))) +
  
  
  labs(x = "",
       y = "Number of Drugs", 
       title = "Total Number of drugs evaluated by the \n European Medicines Agency (1995-2022)",
       subtitle = "Orphan drugs are 12% of drugs approved while authorisation \n rates are around 90% for orphan drugs and regular drugs.",
       # caption = "Data source: European Medicines Agency"
  ) +
  
  theme(
    axis.title.x = element_blank()
  )

orphan_total_plot
```

## Orphan Drugs plot

The combined visualization effectively tells the story of orphan drug development in Europe, from policy implementation to current trends, while maintaining perspective on their proportion within overall drug development.

```{r}
Orphan_drugs_plot <- wrap_plots(orphan_years_plot, orphan_total_plot, ncol = 2, nrow = 1, widths = c(1, 0.33)) + 
  plot_annotation(title = 'Orphan drugs submitted for evaluation by the European Medicines Agency',
                  subtitle = "",
                  caption = "Data source: European Medicines Agency",
                  
                  # tag_levels = 'A', tag_prefix = "Plot ",
                  theme = theme(plot.title = element_text(size = 32)))

Orphan_drugs_plot

ggsave("Orphan drugs plot.png", Orphan_drugs_plot, path="Plots", width = 5400, height = 3000, units = "px", bg="white")
```
