---
title: "2019-01-15 - Space Launches"
format: html
---

# *Introduction*

*These are the data behind the "space launches" article, [The space race is dominated by new contenders](https://economist.com/graphic-detail/2018/10/18/the-space-race-is-dominated-by-new-contenders).*

*Principal data came from the Jonathan McDowell's JSR Launch Vehicle Database, available online at <http://www.planet4589.org/space/lvdb/index.html>.*

## *Launches (launches.csv)*

| *variable* | *definition* |
|------------------------------------|------------------------------------|
| *tag* | *Harvard or [COSPAR](https://en.wikipedia.org/wiki/International_Designator) id of launch* |
| *JD* | [*Julian Date*](https://en.wikipedia.org/wiki/Julian_day) *of launch* |
| *launch_date* | *date of launch* |
| *launch_year* | *year of launch* |
| *type* | *type of launch vehicle* |
| *variant* | *variant of launch vehicle* |
| *mission* |  |
| *agency* | *launching agency* |
| *state_code* | *launching agency's state* |
| *category* | *success (O) or failure (F)* |
| *agency_type* | *type of agency* |

## *Agencies (Agencies.csv)*

| *variable*           | *definition*              |
|----------------------|---------------------------|
| *agency*             | *org phase code*          |
| *count*              | *number of launches*      |
| *ucode*              | *org Ucode*               |
| *state_code*         | *responsible state*       |
| *type*               | *type of org*             |
| *class*              | *class of org*            |
| *tstart*             | *org/phase founding date* |
| *tstop*              | *org/phase ending date*   |
| *short_name*         | *short name*              |
| *name*               | *full name*               |
| *location*           | *plain english location*  |
| *longitude*          |                           |
| *latitude*           |                           |
| *error*              | *uncertainty in long/lat* |
| *parent*             | *parent org*              |
| *short_english_name* | *english short name*      |
| *english_name*       | *english full name*       |
| *unicode_name*       | *unicode full name*       |
| *agency_type*        | *type of agency*          |

# *Data Preparation*

```{r}
#| label: Import libraries
#| warning: False
#| message: False

library(DT)
library(ggimage)
library(tidyverse)
library(patchwork)
library(extrafont)
```

```{r}
#| label: Import Data
#| warning: False
#| message: False

agencies <- read_csv("agencies.csv")
launches <- read_csv("launches.csv")

country_names <- c("BR" = "Brazil", "CN" = "China", "CYM" = "Sea Launch", "F" = "France", "I" = "Italy", "I-ELDO" = "Italy", "I-ESA" = "Italy", "IL" = "Israel", "IN" = "India", "IR" = "Iran", "J" = "Japan",  "KP" = "North Korea", "KR" = "South Korea", "RU" = "Soviet Union - Russia", "SU" = "Soviet Union - Russia", "UK" = "United Kingdom", "US" = "United States"
)

launches <- launches %>% 
  mutate(country_name = country_names[state_code])

launches %>% sample_n(500) %>% arrange(launch_date) %>% datatable(options = list(pageLength = 10, filter = 'top', searching=T))
```

```{r}
#| label: Theme setting
#| warning: False
#| message: False

loadfonts(device = "win")

theme_set(theme_minimal())
theme_update(
  
  plot.title = element_text(
    family = "DejaVu Math TeX Gyre",
    face = "bold",
    size = 32,           
    colour = "black",        
    angle = 0,
    hjust = 0.5,
    vjust = 0.5,
    lineheight = 1.2, 
    margin = margin(t = 10), 
  )
  
)

```

# *Visualizing Launches*

## *Launches by country*

```{r}
#| label: Launches by country
#| collapse: true
#| fig.height: 9
#| fig.width: 16

countries_colors <- c("United States" = "#1a80bb",
                      "France" = "#8cc5e3",
                      "Soviet Union - Russia" = "#a00000",
                      "China" = "#d8a6a6",
                      "Other Countries" = "grey40")

launches_by_country <- launches %>% 
  group_by(country_name) %>% 
  summarise(count = n())

total_country <- nrow(launches_by_country)
keep_n <- 4  # Number of top countries to show in main plot

# Create base theme for shared styling
theme_space_launches <- function() {
  theme_minimal() %+replace%
    theme(  # Shared elements between both plots
      legend.position = "none",
      panel.grid.minor = element_blank(),
      panel.grid.major.y = element_blank(),
      axis.title = element_blank(),
      plot.background = element_rect(fill = "#f8f8f8"),
      plot.title = element_text(face = "bold", margin = margin(b = 10), hjust = 0.5))
}


# Main plot
launches_by_country_top_plot <- launches_by_country %>% 
  mutate(country_name = fct_lump_n(country_name, n = keep_n, w = count, other_level = "Other Countries"),  # Combine all countries beyond top 4 into "Other Countries"
         country_name = fct_reorder(country_name, count)) %>% 
  
  ggplot(aes(x = country_name, y = count, fill = country_name)) +
  geom_col(width = 0.7) +
  coord_flip() +
  
  # Add country flags next to bars
  geom_image(aes(image = paste0("flags/", country_name, ".png")), y = -100, size = 0.08) +
  
  scale_y_continuous(position = "right",
                     labels = scales::comma,
                     breaks = seq(0, 3000, by = 500)) +
  
  scale_fill_manual(values = countries_colors) + 
  
  # Add visual connectors to inset plot
  geom_rect(xmin = 0.5, xmax = 1.5, ymin = 0, ymax = 300, color = "#333333", fill = NA, size = 0.7, linetype = "dashed") +  # Rectangle around "Other Countries" bar
  
  geom_segment(x = 0.5, y = 300, xend = 0.4, yend = 890, color = "#333333", linetype = "dashed") + 
  geom_segment(x = 1.5, y = 300, xend = 3, yend = 890, color = "#333333", linetype = "dashed") + 
  
  labs(title = "Global Space Race: Launch Statistics by Country",
       subtitle = "Total number of orbital launches from 1957 to 2018",
       caption = "Data source: Jonathan McDowell's JSR Launch Vehicle Database") +
  
  # Theme
  theme_space_launches() +
  theme(panel.background = element_rect(fill = "#f8f8f8", color = NA),
        axis.text = element_text(size = 24, color = "#333333"),
        axis.text.x = element_text(margin = margin(t = 10)),
        plot.title = element_text(size = 32),
        plot.subtitle = element_text(size = 24, color = "#666666", margin = margin(b = 20), hjust = 0.5),
        plot.caption = element_text(size = 16, color = "#666666", margin = margin(t = 20)),
        plot.margin = margin(80, 20, 20, 20))

# Inset plot with other countries
launches_by_country_inset_plot <- launches_by_country %>% 
  arrange(count) %>% 
  head(total_country - keep_n) %>% 
  mutate(country_name = fct_reorder(country_name, count)) %>% 
  ggplot(aes(x = country_name, y = count, fill = country_name)) + 
  geom_col(width = 0.7) + 
  scale_y_continuous(position = "right", 
                     breaks = seq(0, 120, by = 20),
                     labels = scales::comma) +
  scale_fill_viridis_d(option = "mako", begin = 0.7, end = 0.3) +
  coord_flip() +
  labs(title = "Other Countries' Launch Statistics",
       caption = "Note: Sea Launch program (1999-2014) involved multiple countries") +
  theme_space_launches() +
  theme(plot.background = element_rect(fill = "#f8f8f8", color = "#333333", size = 0.5),
        plot.margin = unit(c(10, 15, 10, 10), "pt"),
        axis.text = element_text(size = 14, color = "#333333"),
        plot.title = element_text(size = 16),
        plot.caption = element_text(size = 12, color = "#666666", margin = margin(t = 10)))

# Combine main plot with inset plot
launches_by_country_top_plot + 
  inset_element(launches_by_country_inset_plot, 
                l = 0.3, r = 1.0, t = 0.5, b = 0, clip = FALSE)

```

## *The fall of the Soviet Union*

```{r}
#| label: The Fall of the Soviet Union



```
