---
title: "2019-01-15 - Space Launches"
format: html
---

# *Introduction*

*These are the data behind the "space launches" article, [The space race is dominated by new contenders](https://economist.com/graphic-detail/2018/10/18/the-space-race-is-dominated-by-new-contenders).*

*Principal data came from the Jonathan McDowell's JSR Launch Vehicle Database, available online at <http://www.planet4589.org/space/lvdb/index.html>.*

## *Launches (launches.csv)*

| *variable* | *definition* |
|------------------------------------|------------------------------------|
| *tag* | *Harvard or [COSPAR](https://en.wikipedia.org/wiki/International_Designator) id of launch* |
| *JD* | [*Julian Date*](https://en.wikipedia.org/wiki/Julian_day) *of launch* |
| *launch_date* | *date of launch* |
| *launch_year* | *year of launch* |
| *type* | *type of launch vehicle* |
| *variant* | *variant of launch vehicle* |
| *mission* |  |
| *agency* | *launching agency* |
| *state_code* | *launching agency's state* |
| *category* | *success (O) or failure (F)* |
| *agency_type* | *type of agency* |

## *Agencies (Agencies.csv)*

| *variable*           | *definition*              |
|----------------------|---------------------------|
| *agency*             | *org phase code*          |
| *count*              | *number of launches*      |
| *ucode*              | *org Ucode*               |
| *state_code*         | *responsible state*       |
| *type*               | *type of org*             |
| *class*              | *class of org*            |
| *tstart*             | *org/phase founding date* |
| *tstop*              | *org/phase ending date*   |
| *short_name*         | *short name*              |
| *name*               | *full name*               |
| *location*           | *plain english location*  |
| *longitude*          |                           |
| *latitude*           |                           |
| *error*              | *uncertainty in long/lat* |
| *parent*             | *parent org*              |
| *short_english_name* | *english short name*      |
| *english_name*       | *english full name*       |
| *unicode_name*       | *unicode full name*       |
| *agency_type*        | *type of agency*          |

# *Data Preparation*

```{r}
#| label: Import libraries
#| warning: False
#| message: False

library(DT)
library(scales)
library(ggimage)
library(tidyverse)
library(patchwork)
library(extrafont)
```

Let's explore this data

```{r}
#| label: Import Data
#| warning: False
#| message: False

agencies <- read_csv("agencies.csv")
launches <- read_csv("launches.csv")

country_names <- c("BR" = "Brazil", "CN" = "China", "CYM" = "Sea Launch", "F" = "France", "I" = "Italy", "I-ELDO" = "Italy", "I-ESA" = "Italy", "IL" = "Israel", "IN" = "India", "IR" = "Iran", "J" = "Japan",  "KP" = "North Korea", "KR" = "South Korea", "RU" = "Soviet Union - Russia", "SU" = "Soviet Union - Russia", "UK" = "United Kingdom", "US" = "United States"
)

launches <- launches %>% 
  mutate(country_name = country_names[state_code])

launches %>% sample_n(500) %>% arrange(launch_date) %>% datatable(options = list(pageLength = 10, filter = 'top', searching=T))
```

```{r}
#| label: Theme setting
#| warning: False
#| message: False

loadfonts(device = "win")

theme_set(theme_minimal())
theme_update(
  
  plot.title = element_text(
    family = "DejaVu Math TeX Gyre",
    face = "bold",
    size = 32,           
    colour = "black",        
    angle = 0,
    hjust = 0.5,
    vjust = 0.5,
    lineheight = 1.2, 
    margin = margin(t = 10), 
  )
  
)

```

# The Space Race: A Half-Century of Orbital Ambitions

Since the launch of Sputnik 1 in 1957, the quest to reach orbit has been pursued by nations around the world, creating a fascinating story of technological achievement and geopolitical rivalry.

## Global Leaders in Space

China and France follow as the next most active spacefaring nations, though with significantly fewer launches. This hierarchy reflects both technological capabilities and national priorities in space exploration.

```{r}
#| label: Launches by country
#| collapse: true
#| fig.height: 9
#| fig.width: 16

# Define base text size for relative scaling
base_size <- 12

# Create base theme with relative scaling
theme_space_launches <- function(base_size = 12) {
  theme_minimal(base_size = base_size) %+replace%
    theme(
      legend.position = "none",
      panel.grid.minor = element_blank(),
      panel.grid.major.y = element_blank(),
      axis.title = element_blank(),
      plot.background = element_rect(fill = "#f8f8f8"),
      plot.title = element_text(
        face = "bold", 
        margin = margin(b = 10), 
        hjust = 0.5,
        size = rel(2.0)  # Relative to base size
      )
    )
}

# Main plot
launches_by_country_top_plot <- launches_by_country %>% 
  mutate(
    country_name = fct_lump_n(country_name, n = keep_n, w = count, 
                             other_level = "Other Countries"),
    country_name = fct_reorder(country_name, count)
  ) %>% 
  ggplot(aes(x = country_name, y = count, fill = country_name)) +
  geom_col(width = 0.7) +
  coord_flip() +
  
  # Add country flags
  geom_image(aes(image = paste0("flags/", country_name, ".png")), 
             y = -100, size = 0.08) +
  
  scale_y_continuous(
    position = "right",
    labels = scales::comma,
    breaks = seq(0, 3000, by = 500)
  ) +
  
  scale_fill_manual(values = countries_colors) + 
  
  # Inset connectors
  geom_rect(
    xmin = 0.5, xmax = 1.5, ymin = 0, ymax = 300,
    color = "#333333", fill = NA, size = 0.7, linetype = "dashed"
  ) +
  geom_segment(
    x = 0.5, y = 300, xend = 0.4, yend = 890,
    color = "#333333", linetype = "dashed"
  ) + 
  geom_segment(
    x = 1.5, y = 300, xend = 3, yend = 890,
    color = "#333333", linetype = "dashed"
  ) + 
  
  labs(
    title = "Global Space Race: Launch Statistics by Country",
    subtitle = "Total number of orbital launches from 1957 to 2018",
    caption = "Data source: Jonathan McDowell's JSR Launch Vehicle Database"
  ) +
  
  theme_space_launches(base_size) +
  theme(
    panel.background = element_rect(fill = "#f8f8f8", color = NA),
    axis.text = element_text(
      size = rel(2.0),  # 24/12
      color = "#333333"
    ),
    axis.text.x = element_text(margin = margin(t = 10)),
    plot.title = element_text(size = rel(2.67)),  # 32/12
    plot.subtitle = element_text(
      size = rel(2.0),  # 24/12
      color = "#666666",
      margin = margin(b = 20),
      hjust = 0.5
    ),
    plot.caption = element_text(
      size = rel(1.33),  # 16/12
      color = "#666666",
      margin = margin(t = 20)
    ),
    plot.margin = margin(80, 20, 20, 20)
  )

# Inset plot
launches_by_country_inset_plot <- launches_by_country %>% 
  arrange(count) %>% 
  head(total_country - keep_n) %>% 
  mutate(country_name = fct_reorder(country_name, count)) %>% 
  ggplot(aes(x = country_name, y = count, fill = country_name)) + 
  geom_col(width = 0.7) + 
  scale_y_continuous(
    position = "right", 
    breaks = seq(0, 120, by = 20),
    labels = scales::comma
  ) +
  scale_fill_viridis_d(option = "mako", begin = 0.7, end = 0.3) +
  coord_flip() +
  labs(
    title = "Other Countries' Launch Statistics",
    caption = "Note: Sea Launch program (1999-2014) involved multiple countries"
  ) +
  theme_space_launches(base_size) +
  theme(
    plot.background = element_rect(
      fill = "#f8f8f8",
      color = "#333333",
      size = 0.5
    ),
    plot.margin = unit(c(10, 15, 10, 10), "pt"),
    axis.text = element_text(
      size = rel(1.17),  # 14/12
      color = "#333333"
    ),
    plot.title = element_text(size = rel(1.33)),  # 16/12
    plot.caption = element_text(
      size = rel(1.0),  # 12/12
      color = "#666666",
      margin = margin(t = 10)
    )
  )

# Combine plots
launches_by_country_top_plot + 
  inset_element(
    launches_by_country_inset_plot, 
    l = 0.3, r = 1.0, t = 0.5, b = 0, 
    clip = FALSE
  )

```

Our visualization reveals the dominant players in orbital launches from 1957 to 2018. The Soviet Union (later Russia) and the United States emerge as the clear frontrunners, reflecting their historic Space Race during the Cold War. Together, these two nations account for the majority of all orbital launches. They each have more than the combined total of all the other countries in the world.

China and France follow as the next most active spacefaring nations, though with significantly fewer launches. This hierarchy reflects both technological capabilities and national priorities in space exploration. Once again the rest of the nations have a combined total of launches smaller than France.

A diverse group of nations that have developed independent space launch capabilities, including:

-   India, which has steadily built its space program since the 1960s
-   Japan, known for its reliable and sophisticated launch vehicles
-   Israel, which has developed specialized launch capabilities
-   Iran and North Korea, which have more recently joined the ranks of spacefaring nations

Of particular interest is the Sea Launch program, a unique multinational commercial venture that operated from 1999 to 2014. This project, involving Norway, Russia, Ukraine, and the United States, exemplified international cooperation in space access, using a mobile sea-based platform for equatorial launches.

## *The fall of the Soviet Union*

```{r}
#| label: The Fall of the Soviet Union

# Define base text size for relative scaling
base_size <- 12

# Define color scheme consistent with previous plot
countries_colors <- c(
  "United States" = "#1a80bb",
  "France" = "#8cc5e3",
  "Soviet Union - Russia" = "#a00000",
  "China" = "#d8a6a6",
  "Other Countries" = "grey40"
)

# Create base theme with relative scaling
theme_space_timeline <- function(base_size = 12) {
  theme_minimal(base_size = base_size) %+replace%
    theme(
      # Text elements
      plot.title = element_text(
        size = rel(1.83),  # 22/12 - reduced from 2.67
        face = "bold",
        margin = margin(t = 20, b = 10),
        hjust = 0.5
      ),
      plot.subtitle = element_text(
        size = rel(1.33),   # 16/12 - reduced from 2.0
        color = "#666666",
        margin = margin(b = 20),
        hjust = 0.5
      ),
      plot.caption = element_text(
        size = rel(1.33),  # 16/12
        color = "#666666",
        margin = margin(t = 20)
      ),
      
      # Axis formatting
      axis.title = element_text(
        size = rel(1.5),   # 18/12
        color = "#333333"
      ),
      axis.text = element_text(
        size = rel(1.17),  # 14/12
        color = "#333333"
      ),
      
      # Grid lines
      panel.grid.minor = element_blank(),
      panel.grid.major.x = element_line(color = "grey90"),
      panel.grid.major.y = element_line(color = "grey90"),
      
      # Legend formatting
      legend.position = "right",
      legend.title = element_text(size = rel(1.33)),  # 16/12
      legend.text = element_text(size = rel(1.17)),   # 14/12
      legend.key.size = unit(2, "lines"),
      
      # Background
      plot.background = element_rect(fill = "#f8f8f8", color = NA),
      panel.background = element_rect(fill = "#f8f8f8", color = NA),
      
      # Margins
      plot.margin = margin(30, 30, 30, 30)  # Increased overall margins
    )
}

# Calculate cumulative launches per country over time
cumulative_launches <- launches %>%
  arrange(launch_date) %>%
  group_by(country_name) %>%
  mutate(cumulative_count = row_number()) %>%
  mutate(launch_year = year(launch_date)) %>% 
  filter(launch_year <= 2024)

# Create the line plot
cumulative_launches %>% 
  ggplot(aes(x = launch_date, 
             y = cumulative_count, 
             group = country_name)) +
  # Add lines
  geom_line() +
  
  # Customize scales
  scale_y_continuous(
    labels = scales::comma,
    breaks = seq(0, 3000, by = 500),
    expand = expansion(mult = c(0.02, 0.02))
  ) +
  
  scale_x_date(
    date_breaks = "10 years",
    date_labels = "%Y",
    expand = expansion(mult = c(0.02, 0.02))
  ) +
  
  # Add labels
  labs(
    title = "The Space Race: Cumulative Orbital Launches (1957-2018)",
    subtitle = "Showing the progression of space launches by major space-faring nations",
    x = "Year",
    y = "Cumulative Number of Launches",
    caption = "Data source: Jonathan McDowell's JSR Launch Vehicle Database"
  ) +
  
  # Apply themed styling with relative text sizes
  theme_space_timeline(base_size)

```
