---
title: "Bob Ross Paintings"
author: "Hippolyte Menou"
format: html
editor: visual
---

```{r}
# devtools::install_github("frankiethull/BobRossColors")
# devtools::install_github("ctesta01/ggtriangles")

library(patchwork)
library(tidyverse)
library(colorspace)
library(BobRossColors)
library(ggtriangles)
```

```{r}
bob_ross <- read_csv(
  "https://raw.githubusercontent.com/jwilber/Bob_Ross_Paintings/master/data/bob_ross_paintings.csv") %>% 
  select(-1) %>% 
  mutate(across(Black_Gesso:Alizarin_Crimson, as.logical))

bob_ross %>% 
  glimpse()
```

```{r Set Theme}
theme_set(theme_minimal())

theme_update(
  
  plot.background = element_rect(fill = "lightblue", color = "lightblue"),
  panel.background = element_rect(fill = "lightyellow", color = "lightyellow"),
  
  panel.grid.major=element_blank(),
  panel.grid.minor = element_blank(),
  
  axis.title = element_blank(),
  axis.text = element_blank(),
  
  legend.position = "none"
)

```

# How many paintings per season?

There are 13 paintings per season every season for 31 season

```{r}
bob_ross %>% 
  group_by(season) %>% 
  summarise(paintings = n()) %>% 
  arrange(desc(paintings))
```

# How many colors does Bob Ross uses per painting ?

```{r color_number_plot}
bob_ross %>% 
  summarise(median=median(num_colors))

bob_ross %>% 
  count(num_colors, sort=T) %>% 
  summarise(mean = weighted.mean(num_colors, n))

bob_ross %>% 
  count(num_colors, sort=T) %>% 
  arrange(num_colors) %>% 
  add_tally(n, name="total_paintings") %>% 
  mutate(sum_colors = cumsum(n),
         perc = round(sum_colors / total_paintings, 2)) 

bob_ross %>% 
  summarise(interval = list(quantile(num_colors, probs = c(0.05, 0.1, 0.9, 0.95)))) %>%
  unnest_wider(interval)

color_numbers <- bob_ross %>% 
  left_join(bob_ross %>% count(num_colors), by="num_colors")

indian_red <- "#CD5C5C"

color_number_plot <- color_numbers %>% 
  ggplot(aes(x=num_colors)) + 
  geom_boxplot(aes(y=120), width=10, outliers=F, staplewidth =0.5, fill=indian_red) + 
  
  geom_bar(stat = "count", fill=indian_red) + 
  
  scale_x_continuous(breaks = 1:15) +
  scale_y_continuous(breaks = seq(0, 100, 20), expand = c(0, 0.1), limits = c(0,130)) +
  
  labs(title="One Color or Many? How Many Colors Make a Masterpiece", 
       x = "Number of colours used by Bob Ross in his paintings", 
       y ="") +
  
  theme(
    plot.title = element_text(hjust=0.5, size=24),
    axis.text = element_text(),
    axis.title = element_text(),
    panel.grid.major.y = element_line(),
    plot.margin = margin(t = 20, r = 10, b = 10, l = 10)
  )

color_number_plot
```

Bob Ross used on average more than 10 colors in his paintings (mean = 10.7 with a median = 11) and up to 15 colors in 7 of them. In 90% of his paintings, between 6 and 13 colors are used.

Bob Ross used only one color of paint in this painting : "Contemplative Lady" was composed using only Van Dyke Brown which he declined in multiple shades.

```{r One color painting}
bob_ross %>%
  filter(num_colors==1)
# Contemplative Lady

show_colors_and_paintings(paintings = c("contemplative_lady"))
```

# What are the most frequent colors used by Bob Ross

```{r most_frequent_colors_plot}
most_frequent_used_colours <- bob_ross %>% 
  summarise(across(Black_Gesso:Alizarin_Crimson, sum)) %>% 
  pivot_longer(Black_Gesso:Alizarin_Crimson, names_to = "color", values_to = "count") %>% 
  arrange(desc(count)) %>% 
  mutate(color = fct_reorder(color, count)) %>% 
  mutate(color = str_replace_all(color, "_", "")) %>% 
  left_join(unique_bob_ross_colors)

color_palette <- most_frequent_used_colours$color_hex
names(color_palette) <- most_frequent_used_colours$color
color_palette


most_frequent_colors_plot <- most_frequent_used_colours %>% 
  ggplot(aes(x=fct_reorder(color, count), y=count, fill=color)) + 
  geom_col() + 
  coord_flip() +
  ylim(0, 403) +
  scale_y_continuous(limits=c(0,410), expand = c(0,0)) +
  
  scale_fill_manual(values=color_palette) + 
  labs(title="Bob Ross's Favorite Hues", 
       # subtitle="Starting with Titanium White",
       x = NA, y = NA) +
  theme(
    plot.title = element_text(hjust=0.5, size=24),
    plot.subtitle = element_text(hjust=0.5, size=16),

    axis.text = element_text(),
    axis.ticks = element_line(),
    panel.grid.major.x = element_line(),
    plot.margin = margin(t = 20, r = 10, b = 10, l = 10)
  )

most_frequent_colors_plot
```

```{r legend_plot}

get_luminance <- function(color) {
  rgb_color <- col2rgb(color) / 255
  0.2126 * rgb_color[1,] + 0.7152 * rgb_color[2,] + 0.0722 * rgb_color[3,]
}

legend_colors <- most_frequent_used_colours %>% 
  mutate(
    label = color %>%
      str_split("(?=[A-Z])") %>% 
      map_chr(~ str_c(.x, collapse = " ")) %>% 
      str_c(color_hex, sep="\n"),
    
    text_color = if_else(get_luminance(color_hex) > 0.5, "black", "white")) 
legend_colors

legend_plot <- legend_colors %>% 
  ggplot(aes(y = fct_reorder(color, count), x=0)) + 
  geom_tile(aes(fill=color)) + 
  coord_fixed(ratio = 1) + 
  
  geom_text(aes(label=label, color = text_color)) +
  
  scale_color_identity() +
  scale_fill_manual(values=color_palette) 

legend_plot
```

```{r all_seasons_plot}

facet_title <- str_c("Season ", 1:31)
facet_title[-c(1,6,11,16,21,26,31)] <- ""  # Only write the season title for these seasons
names(facet_title) <- 1:31
facet_title

all_seasons_colors <- bob_ross %>% 
  pivot_longer(Black_Gesso:Alizarin_Crimson, names_to = "color") %>% 
  mutate(color = str_replace_all(color, "_", "")) %>% 
  filter(value) %>% 
  left_join(most_frequent_used_colours, by="color")

all_seasons_colors

all_seasons_plot <- all_seasons_colors %>% 
  ggplot(aes(x=episode, y=fct_reorder(color, count))) + 
  geom_tile(aes(color=color, fill=color), width = 1, height = 1) + 
  coord_fixed(ratio = 1) +
  facet_wrap(~season, ncol=5, labeller = labeller(season = facet_title), strip.position = "left") + 
  
  scale_color_manual(values=color_palette) + 
  scale_fill_manual(values=color_palette) + 
  
  theme(
    
    plot.background = element_rect(fill = "lightblue"),
    panel.background = element_rect(fill = "lightyellow"),
    
    panel.grid.major=element_blank(),
    panel.grid.minor = element_blank(),
    
    axis.title = element_blank(),
    axis.text = element_blank())

all_seasons_plot
```

```{r Plot Patchwork}
total_width <- 4500 + 4000
legend_width <- 1000
legend_ratio <- legend_width / total_width
all_ratio <- 1 - legend_ratio


design <- "
  123
  124
  12#
"

bob_ross_colors <- all_seasons_plot + legend_plot + color_number_plot + most_frequent_colors_plot +
  plot_layout(design = design, widths = c(all_ratio, legend_ratio, all_ratio)) + 
  plot_annotation(title = "The Joy of Color: Bob Ross's Happy Little Palette",
                  subtitle = "",
                  caption = "Data source: Jared Wilber's data on Bob Ross Paintings via @frankiethull Bob Ross Colors data package",
                  
                  theme = theme(plot.title = element_text(size = 32, hjust = 0.5)))

bob_ross_colors

ggsave("bob_ross_colors.png", bob_ross_colors, path="Plots", width = total_width, height = 7000, units = "px", bg="white")
```

# Which colors are used together

```{r}
row_occurance <- bob_ross %>% 
  select(painting_index, Black_Gesso:Alizarin_Crimson) %>%
  pivot_longer(Black_Gesso:Alizarin_Crimson, names_to = "color", values_to = "n")


co_occurence <- row_occurance %>%
  left_join(row_occurance, by = "painting_index") %>%
  mutate(occurence = if_else(n.x & n.y, 1, 0)) %>% 
  group_by(color.x, color.y) %>%
  summarise(Occurance = sum(occurence)) %>%
  spread(key = "color.y", value = "Occurance", fill = 0) %>% 
  rename(Color = "color.x")

co_occurence

colors_used_together <- co_occurence %>% 
  pivot_longer(Alizarin_Crimson:Yellow_Ochre, names_to = "Pair", values_to = "n") %>% 
  mutate(Color = str_replace_all(Color, "_", ""), Pair = str_replace_all(Pair, "_", ""))

colors_used_together %>% 
  arrange(-n)

used_together_plot <- colors_used_together %>% 
  ggplot(aes(x=fct_reorder(Color, -n), y=fct_reorder(Pair, n))) +
  geom_point(shape=21, aes(colour=Color, fill=Pair, size=n/400), stroke=1.5) +
  
  coord_fixed(ratio = 1) +

  scale_fill_manual(values=color_palette) + 
  scale_colour_manual(values=color_palette) + 
  
  theme()

used_together_plot
```
